---
title: '"Where is the ... key?"'
knitr: 
  opts_chunk: 
    echo: false
    message: false
    warning: false
    crop: true
---

```{r}
library(tidyverse)
library(ggkeyboard)
knitr::knit_hooks$set(crop = knitr::hook_pdfcrop)
```

## Single Quotes vs Backticks

It's important to distinguish between single quotes: `'` and backticks <code>\`</code>. They look similar, but aren't interchangeable.

The backtick key is highlighted in the images of the keyboards below, in the top left side.

::: panel-tabset
## PC

```{r}
#| fig-width: 10
#| fig-height: 4
#| crop: true
 ggkeyboard(palette = keyboard_palette("wahtsy")) %>%
   highlight_keys(c("~`"),
                  fill = "red", colour = "red",
                  alpha = 0.3)
```

## Mac

```{r}
#| fig-width: 10
#| fig-height: 4
#| crop: true
ggkeyboard(keyboard = mac,
           palette = keyboard_palette("magic")) %>%
   highlight_keys(c("~`"),
                  fill = "red", colour = "red",
                  alpha = 0.3)
```
:::

## Slash (`/`) vs backslash (`\`).

These different symbols will do different things, and aren't interchangeable!

### Slash (`/`) location

::: panel-tabset
## PC

```{r}
#| fig-width: 10
#| fig-height: 4
ggkeyboard(palette = keyboard_palette("wahtsy")) %>%
   highlight_keys(c("?/"),
                  fill = "red", colour = "red",
                  alpha = 0.3)
```

## Mac

```{r}
#| fig-width: 10
#| fig-height: 4
ggkeyboard(mac,
           palette = keyboard_palette("magic")) %>%
   highlight_keys(c("?/"),
                  fill = "red", colour = "red",
                  alpha = 0.3)
```
:::

### Backslash (`\`) location

::: panel-tabset
## PC

```{r}
#| fig-width: 10
#| fig-height: 4
ggkeyboard(palette = keyboard_palette("wahtsy")) %>%
   highlight_keys(c("|\\"),
                  fill = "red", colour = "red",
                  alpha = 0.3)
```

## Mac

```{r}
#| fig-width: 10
#| fig-height: 4
ggkeyboard(mac,
           palette = keyboard_palette("magic")) %>%
   highlight_keys(c("|\\"),
                  fill = "red", colour = "red",
                  alpha = 0.3)
```
:::

## The tilde `~` key.

To type the tilde[^1] key you have to hold {{< kbd Shift+` >}}.

[^1]: Pronunciations include \[tɪl.də\] and \[tɪl.di\]

::: panel-tabset
## PC

```{r}
#| fig-width: 10
#| fig-height: 4
ggkeyboard(keyboard =tkl,
           palette = keyboard_palette("wahtsy")) %>%
   highlight_keys(c("Shift Left", "~`"),
                  fill = "red", colour = "red",
                  alpha = 0.3)
```

## Mac

```{r}
#| fig-width: 10
#| fig-height: 4
ggkeyboard(keyboard = mac,
           palette = keyboard_palette("magic")) %>%
   highlight_keys(c("Shift Left", "~`"),
                  fill = "red", colour = "red",
                  alpha = 0.3)
```
:::

## The caret `^`

You can get the caret symbol with {{< kbd Shift+6 >}}

::: panel-tabset
## PC

```{r}
#| fig-width: 10
#| fig-height: 4
ggkeyboard(keyboard =tkl,
           palette = keyboard_palette("wahtsy")) %>%
   highlight_keys(c("Shift Left", "6"),
                  fill = "red", colour = "red",
                  alpha = 0.3)
```

## Mac

```{r}
#| fig-width: 10
#| fig-height: 4
ggkeyboard(keyboard = mac,
           palette = keyboard_palette("magic")) %>%
   highlight_keys(c("Shift Left", "6"),
                  fill = "red", colour = "red",
                  alpha = 0.3)
```
:::

## The Pipe `|`

You can get the pipe, or vertical bar, with {{< kbd Shift+\\ >}}.

::: panel-tabset
## PC

```{r}
#| fig-width: 10
#| fig-height: 4
ggkeyboard(keyboard =tkl,
           palette = keyboard_palette("wahtsy")) %>%
   highlight_keys(c("Shift Left", "|\\"),
                  fill = "red", colour = "red",
                  alpha = 0.3)
```

## Mac

```{r}
#| fig-width: 10
#| fig-height: 4
ggkeyboard(keyboard = mac,
           palette = keyboard_palette("magic")) %>%
   highlight_keys(c("Shift Left", "|\\"),
                  fill = "red", colour = "red",
                  alpha = 0.3)
```
:::

## Side Note: "Smart Quotes"

If, at some point, you copy-paste text from the internet into your code or a text file, you might wind up with "smart quotes" that make things work weird!

| Regular Quotes | Smart Quotes |
|----------------|--------------|
| `"regular"`    | `“smart”`      |
| `'regular'`    | `‘smart’`   |

Be careful not to have smart quotes cropping up in your code or data!
